The effect of taxonomic bias on differential-abundance analysis in microbiome studies

### 2022-06-07 differential-abundance.Rmd
Small typo and a couple of notes

The building blocks of a multiplicative DA analysis are the measured FDs in a species' abundance between pairs of individual samples.
<!-- The FE in the measured FD between two samples is given by the ratio of the FEs for the individual-sample measurements. -->
If the fold error (FE) in a species' abundance measurement is constant across samples, then it will not affect the measured FDs between samples.
<!-- Ratios -->
In Section \@ref(abundance-measurement), we showed that consistent taxonomic bias creates a constant FE in the ratio between two species $i$ and $j$, equal to the ratio in their efficiencies (Equation \@ref(eq:ratio-error)).
This error completely cancels when calculating the FD of this ratio from sample $a$ to $b$,
\begin{align}
  (\#eq:ratio-fc-error)
\underbrace{\frac{\tilde R_{i/j}^{(b)}}{\tilde R_{i/j}^{(a)}}}_\text{measured FD}
  = \frac
    {R_{i/j}^{(b)} \cdot \cancel{B_i / B_j}}
    {R_{i/j}^{(a)} \cdot \cancel{B_i / B_j}}
  =
  \underbrace{\frac{R_{i/j}^{(b)}}{R_{i/j}^{(a)}}}_\text{actual FD}
  .
\end{align}
<!-- Proportions -->
In contrast, bias created a FE in the proportion of a species (Equation \@ref(eq:prop-error)) that varied inversely with the mean efficiency of the sample.
Therefore, this error does not cancel when calculating the FD in the proportion,
\begin{align}
  (\#eq:prop-fc-error)
\underbrace{\frac{\tilde P_i^{(b)}}{\tilde P_i^{(a)}}}_\text{measured FD}
  = \frac
    {P_i(b) \cdot \cancel{B_i} / \bar B^{(b)}}
    {P_i(a) \cdot \cancel{B_i} / \bar B^{(a)}}
  =
  \underbrace{\frac{P_i^{(b)}}{P_i^{(a)}}}_\text{actual FD}
  \cdot
  \underbrace{\left[\frac{\bar B^{(b)}}{\bar B^{(a)}}\right]^{-1}}_\text{FE}
  .
\end{align}
The varying mean efficiency does not cancel, leaving an FE in the FD of the species' proportion that is equal to the inverse FD in the mean efficiency.
Notably, this FE is the same for every species.

The different effects of taxonomic bias on the measured FDs in proportions versus ratios are illustrated in a hypothetical example in Figure \@ref(fig:error-proportions) (bottom row).
In this example, the mean efficiency decreases 2.6-fold from Sample 1 to Sample 2 (FD of 0.4), causing the FD in the proportion of each species to appear 2.6-fold larger than its true value.
Though the FE in the FD is the same for each species, the biological implications of the error vary.
In particular, there are three distinct types of error: an increase in the magnitude of the measured FD, a decrease in that magnitude, and a change in the direction of the FD.
We refer to the latter as a _sign error_ in reference to the sign of the measured log fold difference (LFD).
We can see each type of error in Figure \@ref(fig:error-proportions).
For Species 1, which increases in abundance and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FD (actual FD: 2.3, measured FD: 6.5).
For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see a decrease in magnitude (actual FD: 0.15, measured FD: 0.44).
For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FD: 0.6, measured FD: 1.8), such that the species actually appears to increase (a sign error).
In contrast, the ratios between species change by the same factors in the actual and measured taxonomic profiles.
For example, the ratio of Species 2 to Species 3 changes from $1/1$ to $1/4$ in the actual profiles and from $3$ to $3/4$ in the measured profiles, a decrease by 4-fold either way.

>>>>

The building blocks of a multiplicative DA analysis are the measured FDs in a species' abundance between pairs of individual samples.
<!-- The FE in the measured FD between two samples is given by the ratio of the FEs for the individual-sample measurements. -->
If the fold error (FE) in a species' abundance measurement is constant across samples, then it will not affect the measured FDs between samples.
<!-- Ratios -->
In Section \@ref(abundance-measurement), we showed that consistent taxonomic bias creates a constant FE in the ratio between two species $i$ and $j$, equal to the ratio in their efficiencies (Equation \@ref(eq:ratio-error)).
This error completely cancels when calculating the FD of this ratio from sample $a$ to $b$,
\begin{align}
  (\#eq:ratio-fc-error)
\underbrace{\frac{\tilde R_{i/j}^{(b)}}{\tilde R_{i/j}^{(a)}}}_\text{measured FD}
  = \frac
    {R_{i/j}^{(b)} \cdot \cancel{B_i / B_j}}
    {R_{i/j}^{(a)} \cdot \cancel{B_i / B_j}}
  =
  \underbrace{\frac{R_{i/j}^{(b)}}{R_{i/j}^{(a)}}}_\text{actual FD}
  .
\end{align}
<!-- Proportions -->
In contrast, bias created a FE in the proportion of a species (Equation \@ref(eq:prop-error)) that varied inversely with the mean efficiency of the sample.
Therefore, this error does not cancel when calculating the FD in the proportion,
\begin{align}
  (\#eq:prop-fc-error)
\underbrace{\frac{\tilde P_i^{(b)}}{\tilde P_i^{(a)}}}_\text{measured FD}
  = \frac
    {P_i^{(b)} \cdot \cancel{B_i} / \bar B^{(b)}}
    {P_i^{(a)} \cdot \cancel{B_i} / \bar B^{(a)}}
  =
  \underbrace{\frac{P_i^{(b)}}{P_i^{(a)}}}_\text{actual FD}
  \cdot
  \underbrace{\left[\frac{\bar B^{(b)}}{\bar B^{(a)}}\right]^{-1}}_\text{FE}
  .
\end{align}
The varying mean efficiency does not cancel, leaving an FE in the FD of the species' proportion that is equal to the inverse FD in the mean efficiency.
Notably, this FE is the same for every species.

The different effects of taxonomic bias on the measured FDs in proportions versus ratios are illustrated in a hypothetical example in Figure \@ref(fig:error-proportions) (bottom row).
In this example, the mean efficiency decreases 2.6-fold from Sample 1 to Sample 2 (FD of 0.4), causing the FD in the proportion of each species to appear 2.6-fold larger than its true value.
<!-- 0.4 is 1 / 2.6? Is there a reason why 0.4 is the canonical FD rather than 2.6 -->

Though the FE in the FD is the same for each species, the biological implications of the error vary.
In particular, there are three distinct types of error: an increase in the magnitude of the measured FD, a decrease in that magnitude, and a change in the direction of the FD.
We refer to the latter as a _sign error_ in reference to the sign of the measured log fold difference (LFD).
We can see each type of error in Figure \@ref(fig:error-proportions).
For Species 1, which increases in abundance and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FD (actual FD: 2.3, measured FD: 6.5).
For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see a decrease in magnitude (actual FD: 0.15, measured FD: 0.44).
<!-- I was initially confused by calling 0.44 a decrease in magnitude as compared with 0.15 as |0.44| > |0.15|, but see now that 0.15 is a larger multiplicative factor e.g. so that |log(0.44)| < |log(0.15)| -->
For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FD: 0.6, measured FD: 1.8), such that the species actually appears to increase (a sign error).
<!-- Again the sign error is on the log scale (unsure if being more explicit about this is helpful) -->
In contrast, the ratios between species change by the same factors in the actual and measured taxonomic profiles.
For example, the ratio of Species 2 to Species 3 changes from $1/1$ to $1/4$ in the actual profiles and from $3$ to $3/4$ in the measured profiles, a decrease by 4-fold either way.
